Generalized universal instability: Transient linear amplification and subcritical turbulence

### 2015-04-09 universalModeTurbulencePaper.tex
Formatted for JPP

%\documentclass{amsart}
%\documentclass[aip,graphicx]{revtex4-1}
\documentclass[12pt,superscriptaddress]{revtex4}
\usepackage{latexsym, color, graphicx}
%\usepackage{latexsym, color, graphicx, comment}
\usepackage{amssymb}
\usepackage{amsmath}

%\graphicspath{{../MATLAB/}}

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
\newcommand{\erf}{\mathrm{erf}}
\newcommand{\PS}{Pfirsch-Schl\"{u}ter~}
\newcommand{\ExB}{\vect{E}\times\vect{B}}
\newcommand{\Cnl}{C_{\mathrm{nl}a}}
\newcommand{\vma}{\vect{v}_{\mathrm{m}a}}
\newcommand{\vm}{\vect{v}_{\mathrm{m}}}
\newcommand{\vda}{\vect{v}_{\mathrm{d}a}}
\newcommand{\vE}{\vect{v}_{E}}
\newcommand{\vdao}{\vect{v}_{\mathrm{d}a0}}
\newcommand{\vdo}{\vect{v}_{\mathrm{d}0}}
\newcommand{\vEo}{\vect{v}_{E0}}
\newcommand{\vpar}{v_{||}}
\newcommand{\vperp}{v_{\bot}}
\newcommand{\rhop}{\rho_{\theta}}
\newcommand{\rT}{r_{T}}
\newcommand{\Ti}{T_{\mathrm{i}}}
\newcommand{\rTi}{r_{T\mathrm{i}}}
\newcommand{\rTe}{r_{T\mathrm{e}}}
\newcommand{\reta}{r_{\eta}}
\newcommand{\rn}{r_{n}}
\newcommand{\rperp}{r_\bot}
\newcommand{\fM}{f_{\mathrm{M}}}
\newcommand{\fMa}{f_{\mathrm{M}a}}
\newcommand{\fMi}{f_{\mathrm{M}i}}
\newcommand{\fMe}{f_{\mathrm{Me}}}
\newcommand{\fe}{f_{\mathrm{e}}}
\newcommand{\Te}{T_{\mathrm{e}}}
\newcommand{\vth}{v_{\mathrm{th}}}
\newcommand{\nuii}{\nu_{ii}}
\newcommand{\nuPrime}{\hat{\nu}}
\newcommand{\gradpar}{\nabla_{||}}
\newcommand{\kpar}{k_{||}}
\newcommand{\psia}{\psi_0}
\newcommand{\psiaHat}{\hat{\psi}_0}
\newcommand{\omegase}{\omega_{*}}
\newcommand{\nuee}{\nu_{ee}}
\newcommand{\energy}{\mathcal{E}}
\newcommand{\sgn}{\mathrm{sgn}}
\newcommand{\changed}[1]{\textcolor{red}{#1}}
%\newcommand{\changed}[1]{#1}
\newcommand{\transpose}{\mathrm{T}}

\begin{document}

\title{Generalized universal instability: Transient linear amplification and subcritical turbulence}

@@

\author{Matt Landreman}
\email[]{mattland@umd.edu}
\affiliation{Institute for Research in Electronics and Applied Physics, University of Maryland, College Park, MD, 20742, USA}
\author{Gabriel G. Plunk}
\affiliation{Max Planck Institute for Plasma Physics, EURATOM Association and Max-Planck/Princeton Research Center for Plasma Physics}
\author{William Dorland}
\affiliation{Institute for Research in Electronics and Applied Physics, University of Maryland, College Park, MD, 20742, USA}
%\homepage[]{Your web page}
%\thanks{}
%\author{co-authors}

% Collaboration name, if desired (requires use of superscriptaddress option in \documentclass).
% \noaffiliation is required (may also be used with the \author command).
%\collaboration{}
%\noaffiliation

\date{\today}

@@

\pacs{}% insert suggested PACS numbers in braces on next line

\maketitle %\maketitle must follow title, authors, abstract and \pacs

@@

Despite the usefulness of linear eigenmode analysis in many contexts,
it has been established that
eigenmode analysis can in practice give misleading predictions
about the stability of various systems \cite{TrefethenEmbree}.
In particular, turbulence may be sustained in a number of fluid systems
despite the absence of any unstable eigenmodes,
a phenomenon that may be called ``subcritical turbulence''.
The first systems known to display this behavior were certain sheared flow patterns in
neutral fluids, such as plane Poiseuille or plane Couette flow
\cite{Tillmark,TrefethenSubcritical,Grossman}.
Later, sustained turbulence was seen
in fluid simulations of several plasma systems with no absolute linear instabilities
\cite{Waltz, Scott1, Scott2, Drake}.
More recently,
turbulence in the absence of linear instability has been seen in kinetic plasma simulations involving
sheared equilibrium flow \cite{EdmundPRL, BarnesRotation}.
These examples demonstrate that linear eigenmode analysis can give
deceiving predictions for nonlinear behavior,
a conclusion supported by other studies of plasma turbulence in which
linear instability exists, but
the fluctuations in saturated nonlinear turbulence are different in character from the linear
eigenmodes \cite{Biskamp,Friedman1,Friedman2}.

For the cases of turbulence without absolute linear instability in the presence of sheared flows,
for both neutral fluids and plasmas, the phenomenon of transient linear amplification
appears to be important
\cite{TrefethenSubcritical, FarrellIoannou, Grossman, DelSoleSurvey, TrefethenEmbree, Schmid, Newton, Alex}.
Also known as non-modal amplification, transient amplification is a period of growth before
the exponential decay sets in.
A necessary (but not sufficient) condition for transient amplification
is that the linear operator must not have a complete orthogonal set of eigenvectors.
Analysis of transient amplification has also been applied to plasma systems in which
linear instability exists but transient linear behavior may be
more important \cite{Camargo, Camporeale2010, Friedman2, SquirePRL, SquireApJ}.

In the present work, we consider the phenomena of transient amplification and turbulence
without instability for the case of the weakly-collisional
plasma drift wave driven by a density gradient\cite{Galeev,Krall3},
neglecting toroidal effects.
This oscillation has been called the ``universal instability'' as it was at first
thought to be linearly unstable in any plasma with a density gradient.
The oscillation is electrostatic in nature, and linear instability can indeed be found
in slab geometry without magnetic shear.
It was later established that when any amount of magnetic shear is included in the model,
no matter how small,
the eigenmodes with $k_y \rho_i \ll 1$ become linearly damped,
where $k_y$ is the perpendicular wavenumber and $\rho_i$ is the ion gyroradius
\cite{Ross, Tsang, Antonsen}.
However, instability can persist at $k_y \rho_i \ge 1$ when the collisionality is
very low \cite{usUniversalInstability}.
When both magnetic shear and modest collisions are included \cite{usUniversalInstability},
the mode becomes linearly stable at all $k_y \rho_i$.
It is this latter situation that we will consider here.
(As an aside, the small-but-finite-shear eigenmodes
are different from zero-shear modes not only for the
universal instability, but also for the slab ion-temperature-gradient (ITG) instability.
Specifically, in \cite{Plunk2014}, considering the slab mode by setting
curvature drift effects $\omega_d$ to 0, the small-shear limit of (19) is different from
the zero shear result (12).)

Fluid versions of this drift wave system were investigated in \cite{Scott1,Scott2,Drake}.
%using a model with three fluctuating quantities (density, electrostatic potential, and parallel velocity).
In these models, all linear modes were damped, yet nonlinearly, turbulence could be sustained.
In \cite{Drake} it was pointed out that an important secondary
instability in this system is a drift wave rotated by $\pi/2$ in the plane
perpendicular to the magnetic field, since the stabilizing effects of magnetic
shear are reduced in this orientation.
In this rotated drift instability, the role of the equilibrium density gradient is replaced in a sense by the
perturbed density gradient, the latter represented in the nonlinear term.
However, since the nonlinearity is conservative, the rotated drift wave does not inject energy
into the system in the way that can be done by a perturbation feeding off the density gradient of the equilibrium.
Several authors have also investigated transient linear effects in fluid drift wave models
\cite{Camargo, Friedman1, Friedman2}. However, these studies did not
include magnetic shear, which we will find to have a very strong effect on
some aspects of the transient behavior.

In the following sections, we establish three results concerning the drift wave system in slab geometry
with magnetic shear,
extending previous work by considering a kinetic description and exploring the role of transient linear amplification.
For these studies, the initial-value gyrokinetic code gs2 \cite{gs2} will be used
for numerical computations.
First, in section \ref{sec:transient} we demonstrate that
significant transient linear amplification is possible in this system.
The transiently growing fluctuations
resemble the zero-shear eigenmodes
in the dependence of their instantaneous growth rate on wavenumber,
though the transients are sensitive to magnetic shear
in their global time-dependence.
Second, in section \ref{sec:turbulence} we demonstrate that universal-mode turbulence without linear
instability, seen previously in fluid models in \cite{Scott1,Scott2,Drake},
is also possible in the gyrokinetic model.
Finally, we prove there is a relationship between
transient linear amplification and nonlinear turbulence in this system,
using similar arguments to the proof for hydrodynamics in \cite{DelSoleNecessity}
and arriving at similar conclusions.
Specifically, if the physical parameters are such that turbulence can be sustained,
typical states of the turbulence must be amplified linearly for some amount of time, even if these states
eventually decay exponentially.
Two more precise statements of this result and their proof will be given in section \ref{sec:proof}.

@@

We consider slab geometry with magnetic shear, so the magnetic field in Cartesian
coordinates $(x',y',z')$ is $\vect{B} = [\vect{e}_{z'} + (x' / L_s) \vect{e}_{y'} ]B$
for some constant $B$ and shear length $L_s$.  A field-aligned coordinate system
is introduced: $x = x', \;y = y' - x' z' / L_s, \; z = z'$,
so $\vect{B}\cdot\nabla x = \vect{B}\cdot\nabla y = 0$ and $\vect{B}\cdot\nabla z = B$.
In this geometry, we consider the nonlinear electrostatic
gyrokinetic equation \cite{FriemanChen},
dropping toroidal effects:
\begin{eqnarray}
\label{eq:gke}
\frac{\partial h_s}{\partial t}
+ v_{||} \frac{\partial h_s}{\partial z}
+ \frac{1}{B} \left[
\frac{\partial \left<\Phi\right>_{\vect{R}_s}}{\partial X_s} \frac{\partial h_s}{\partial Y_s}
-\frac{\partial \left<\Phi\right>_{\vect{R}_s}}{\partial Y_s} \frac{\partial h_s}{\partial X_s}
\right]
-\left< C_s\left\{h_s\right\} \right>_{\vect{R}_s} \\
=
\frac{q_s}{T_s} f_{Ms} \frac{\partial \left<\Phi\right>_{\vect{R}_s}}{\partial t}
-\frac{f_{Ms}}{B L_n} \frac{\partial \left< \Phi\right>_{\vect{R}_s}}{\partial Y_s}
\left[ 1 + \left( \energy_s-\frac{3}{2}\right) \eta_s\right],
\nonumber
\end{eqnarray}
together with the quasineutrality condition
\begin{equation}
-\frac{n e \Phi}{T_i} + \int d^3v \left<h_i\right>_{\vect{r}}
=
\frac{n e \Phi}{T_e} + \int d^3v \left<h_e\right>_{\vect{r}}.
\label{eq:qn}
\end{equation}
Here $s \in \left\{i,e\right\}$ subscripts denote species,
$\Phi(t,\vect{r})$ is the electrostatic potential, $h_s(t,\vect{R}_s,\energy_s,\mu)$ is the nonadiabatic distribution function,
$\energy_s = m_s v^2/(2 T_s)$, $\mu = v_{\bot}^2/(2B)$,
and $f_{Ms}$ is the leading-order Maxwellian.
The scale length of the equilibrium density $n$ is $L_n = -n/(dn/dx)$,
$q_i = e = -q_e$ is the proton charge,
$T_s$ is the species temperature,
\changed{$\eta_s=(d \ln T_s/dx)/(d \ln n/dx)$,}
$C_s$ is the collision operator,
and
$\left<\right>_{\vect{r}}$
and
$\left<\right>_{\vect{R}_s}$
denote gyroaverages at fixed particle position $\vect{r}=(x,y,z)$ and guiding-center position $\vect{R}_s=(X_s,Y_s,Z_s)$.
The distinction between $z$ and $Z_s$ may be neglected.
Expanding the fluctuating quantities in spatial Fourier modes as $\Phi = \sum_{k_x,k_y} \Phi_{k_x,k_y}(t,z) \exp(i k_x x + i k_y y)$
and $h_s = \sum_{k_x,k_y} h_{s,k_x,k_y}(t,z,\energy_s,\mu) \exp(i k_x X_s + i k_y Y_s)$,
we may replace the gyro-averages with Bessel functions:
$\left<\Phi_{k_x,k_y} \exp(i k_x x + i k_y y)\right>_{\vect{R}_s}
=J_{0s} \Phi_{k_x,k_y} \exp(i k_x X_s + i k_y Y_s)$
and
$\left<h_{s,k_x,k_y}\exp(i k_x X_s + i k_y Y_s)\right>_{\vect{r}} = J_{0s} h_{s,k_x,k_y}\exp(i k_x x + i k_y y)$
with $J_{0s} = J_0(k_{\perp} v_{\perp} / \Omega_s)$
and $\Omega_s = e B/m_s$.
The perpendicular wavenumber is given by
\begin{equation}
k_{\perp} = \sqrt{k_y^2 + (k_x - k_y z/L_s)^2}.
\label{eq:kperp}
\end{equation}
The system is uniquely specified by the following dimensionless parameters:
$L_s / L_n$, $\eta_i$, $\eta_e$, $T_e / T_i$, and collisionality $\nuee L_n / v_i$,
where $\nuee = \sqrt{2}\pi n e^4 \ln\Lambda/[(4\pi\epsilon_0)^2 m_e^{1/2} T_e^{3/2}]$ is the electron-electron collision frequency
and $v_i = \sqrt{2 T_i / m_i}$.

@@

@@

>>>>

% Copyright (C) 2011 Cambridge University Press

\NeedsTeXFormat{LaTeX2e}

\documentclass{jpp}

\usepackage{graphicx}
\usepackage{natbib}

% See if the author has AMS Euler fonts installed: If they have, attempt
% to use the 'upmath' package to provide upright math.
\ifCUPmtlplainloaded \else
  \checkfont{eurm10}
  \iffontfound
    \IfFileExists{upmath.sty}
      {\typeout{^^JFound AMS Euler Roman fonts on the system,
                   using the 'upmath' package.^^J}%
       \usepackage{upmath}}
      {\typeout{^^JFound AMS Euler Roman fonts on the system, but you
                   dont seem to have the}%
       \typeout{'upmath' package installed. JPP.cls can take advantage
                 of these fonts, if you use 'upmath' package.^^J}%
       \providecommand\upi{\pi}%
      }
  \else
    \providecommand\upi{\pi}%
  \fi
\fi

% See if the author has AMS symbol fonts installed: If they have, attempt
% to use the 'amssymb' package to provide the AMS symbol characters.

\ifCUPmtlplainloaded \else
  \checkfont{msam10}
  \iffontfound
    \IfFileExists{amssymb.sty}
      {\typeout{^^JFound AMS Symbol fonts on the system, using the
                'amssymb' package.^^J}%
       \usepackage{amssymb}%
       \let\le=\leqslant  \let\leq=\leqslant
       \let\ge=\geqslant  \let\geq=\geqslant
      }{}
  \fi
\fi

% See if the author has the AMS 'amsbsy' package installed: If they have,
% use it to provide better bold math support (with \boldsymbol).

\ifCUPmtlplainloaded \else
  \IfFileExists{amsbsy.sty}
    {\typeout{^^JFound the 'amsbsy' package on the system, using it.^^J}%
     \usepackage{amsbsy}}
    {\providecommand\boldsymbol[1]{\mbox{\boldmath $##1$}}}
\fi

%%% Example macros (some are not used in this sample file) %%%

% For units of measure
\newcommand\dynpercm{\nobreak\mbox{$\;$dyn\,cm$^{-1}$}}
\newcommand\cmpermin{\nobreak\mbox{$\;$cm\,min$^{-1}$}}

% Various bold symbols
\providecommand\bnabla{\boldsymbol{\nabla}}
\providecommand\bcdot{\boldsymbol{\cdot}}
\newcommand\biS{\boldsymbol{S}}
\newcommand\etb{\boldsymbol{\eta}}

% For multiletter symbols
\newcommand\Real{\mbox{Re}} % cf plain TeX's \Re and Reynolds number
\newcommand\Imag{\mbox{Im}} % cf plain TeX's \Im
\newcommand\Rey{\mbox{\textit{Re}}}  % Reynolds number
\newcommand\Pran{\mbox{\textit{Pr}}} % Prandtl number, cf TeX's \Pr product
\newcommand\Pen{\mbox{\textit{Pe}}}  % Peclet number
\newcommand\Ai{\mbox{Ai}}            % Airy function
\newcommand\Bi{\mbox{Bi}}            % Airy function

% For sans serif characters:
% The following macros are setup in JPP.cls for sans-serif fonts in text
% and math.  If you use these macros in your article, the required fonts
% will be substitued when you article is typeset by the typesetter.
%
% \textsfi, \mathsfi   : sans-serif slanted
% \textsfb, \mathsfb   : sans-serif bold
% \textsfbi, \mathsfbi : sans-serif bold slanted (doesnt exist in CM fonts)
%
% For san-serif roman use \textsf and \mathsf as normal.
%
\newcommand\ssC{\mathsf{C}}    % for sans serif C
\newcommand\sfsP{\mathsfi{P}}  % for sans serif sloping P
\newcommand\slsQ{\mathsfbi{Q}} % for sans serif bold-sloping Q

% Hat position
\newcommand\hatp{\skew3\hat{p}}      % p with hat
\newcommand\hatR{\skew3\hat{R}}      % R with hat
\newcommand\hatRR{\skew3\hat{\hatR}} % R with 2 hats
\newcommand\doubletildesigma{\skew2\tilde{\skew2\tilde{\Sigma}}}
%       italic Sigma with double tilde

% array strut to make delimiters come out right size both ends
\newsavebox{\astrutbox}
\sbox{\astrutbox}{\rule[-5pt]{0pt}{20pt}}
\newcommand{\astrut}{\usebox{\astrutbox}}

\newcommand\GaPQ{\ensuremath{G_a(P,Q)}}
\newcommand\GsPQ{\ensuremath{G_s(P,Q)}}
\newcommand\p{\ensuremath{\partial}}
\newcommand\tti{\ensuremath{\rightarrow\infty}}
\newcommand\kgd{\ensuremath{k\gamma d}}
\newcommand\shalf{\ensuremath{{\scriptstyle\frac{1}{2}}}}
\newcommand\sh{\ensuremath{^{\shalf}}}
\newcommand\smh{\ensuremath{^{-\shalf}}}
\newcommand\squart{\ensuremath{{\textstyle\frac{1}{4}}}}
\newcommand\thalf{\ensuremath{{\textstyle\frac{1}{2}}}}
\newcommand\Gat{\ensuremath{\widetilde{G_a}}}
\newcommand\ttz{\ensuremath{\rightarrow 0}}
\newcommand\ndq{\ensuremath{\frac{\mbox{$\partial$}}{\mbox{$\partial$} n_q}}}
\newcommand\sumjm{\ensuremath{\sum_{j=1}^{M}}}
\newcommand\pvi{\ensuremath{\int_0^{\infty}%
  \mskip \ifCUPmtlplainloaded -30mu\else -33mu\fi -\quad}}

\newcommand\etal{\mbox{\textit{et al.}}}
\newcommand\etc{etc.\ }
\newcommand\eg{e.g.\ }

\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{statement}{Statement}

\title[Generalized universal instability]{Generalized universal instability: Transient linear amplification and subcritical turbulence}

%\author[A. N. Other, H.-C. Smith and J. Q. Public]%
%{A\ls L\ls A\ls N\ns N.\ns O\ls T\ls H\ls E\ls R$^1$%
%  \thanks{Email address for correspondence: plaeditorial@cambridge.org},\ns
%H.\ls-\ls C.\ns S\ls M\ls I\ls T\ls H$^1$\break
%\and J.\ns Q.\ns P\ls U\ls B\ls L\ls I\ls C$^2$}
%
%% NOTE: A full address must be provided: department, university/institution, town/city, zipcode/postcode, country.
%\affiliation{$^1$Plasma Physics Laboratory, University of America,
%Somewhere, IN 12345, USA\\[\affilskip]
%$^2$Department of Physics, University of
%Camford, Academic Street, Camford CF3 5QL, UK}

\author[Matt Landreman, Gabriel G. Plunk and William Dorland]
{M\ls A\ls T\ls T\ns   L\ls A\ls N\ls D\ls R\ls E\ls M\ls A\ls N$^1$
  \thanks{Email address for correspondence: mattland@umd.edu},\ns
G\ls A\ls B\ls R\ls I\ls E\ls L\ns   G.\ns  P\ls L\ls U\ls N\ls K$^2$\break
\and W\ls I\ls L\ls L\ls I\ls A\ls M\ns   D\ls O\ls R\ls L\ls A\ls N\ls D$^1$}

\affiliation{$^1$Institute for Research in Electronics and Applied Physics, University of Maryland, College Park, MD, 20742, USA\\[\affilskip]
$^2$Max Planck Institute for Plasma Physics, Max-Planck/Princeton Research Center for Plasma Physics,
17491 Greifswald, Germany}

\pubyear{2015}
\volume{??}
\pagerange{??}
% Do not enter received and revised dates. These will be entered by the editorial office.
\date{?; revised ?; accepted ?. - To be entered by editorial office}
%\setcounter{page}{1}
\begin{document}

%\documentclass[12pt,superscriptaddress]{revtex4}
%\usepackage{latexsym, color, graphicx}
%\usepackage{amssymb}
%\usepackage{amsmath}

%\newcommand{\red}[1]{\textcolor{red}{#1}}
%\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
\newcommand{\vect}[1]{\boldsymbol{$#1$}}
%\newcommand{\erf}{\mathrm{erf}}
%\newcommand{\PS}{Pfirsch-Schl\"{u}ter~}
%\newcommand{\ExB}{\vect{E}\times\vect{B}}
%\newcommand{\Cnl}{C_{\mathrm{nl}a}}
%\newcommand{\vma}{\vect{v}_{\mathrm{m}a}}
%\newcommand{\vm}{\vect{v}_{\mathrm{m}}}
%\newcommand{\vda}{\vect{v}_{\mathrm{d}a}}
%\newcommand{\vE}{\vect{v}_{E}}
%\newcommand{\vdao}{\vect{v}_{\mathrm{d}a0}}
%\newcommand{\vdo}{\vect{v}_{\mathrm{d}0}}
%\newcommand{\vEo}{\vect{v}_{E0}}
%\newcommand{\vpar}{v_{||}}
%\newcommand{\vperp}{v_{\bot}}
%\newcommand{\rhop}{\rho_{\theta}}
%\newcommand{\rT}{r_{T}}
%\newcommand{\Ti}{T_{\mathrm{i}}}
%\newcommand{\rTi}{r_{T\mathrm{i}}}
%\newcommand{\rTe}{r_{T\mathrm{e}}}
%\newcommand{\reta}{r_{\eta}}
%\newcommand{\rn}{r_{n}}
%\newcommand{\rperp}{r_\bot}
%\newcommand{\fM}{f_{\mathrm{M}}}
%\newcommand{\fMa}{f_{\mathrm{M}a}}
%\newcommand{\fMi}{f_{\mathrm{M}i}}
%\newcommand{\fMe}{f_{\mathrm{Me}}}
%\newcommand{\fe}{f_{\mathrm{e}}}
%\newcommand{\Te}{T_{\mathrm{e}}}
%\newcommand{\vth}{v_{\mathrm{th}}}
%\newcommand{\nuii}{\nu_{ii}}
%\newcommand{\nuPrime}{\hat{\nu}}
%\newcommand{\gradpar}{\nabla_{||}}
%\newcommand{\kpar}{k_{||}}
%\newcommand{\psia}{\psi_0}
%\newcommand{\psiaHat}{\hat{\psi}_0}
\newcommand{\omegase}{\omega_{*}}
\newcommand{\nuee}{\nu_{ee}}
\newcommand{\energy}{\mathcal{E}}
%\newcommand{\sgn}{\mathrm{sgn}}
\newcommand{\sgn}{\mbox{sgn}}
%%\newcommand{\changed}[1]{\textcolor{red}{#1}}
%\newcommand{\changed}[1]{#1}
\newcommand{\transpose}{\mathrm{T}}

@@

\maketitle

@@

\begin{PACS}
???
\end{PACS}

@@

Despite the usefulness of linear eigenmode analysis in many contexts,
it has been established that
eigenmode analysis can in practice give misleading predictions
about the stability of various systems \citep{TrefethenEmbree}.
In particular, turbulence may be sustained in a number of fluid systems
despite the absence of any unstable eigenmodes,
a phenomenon that may be called ``subcritical turbulence''.
The first systems known to display this behaviour were certain sheared flow patterns in
neutral fluids, such as plane Poiseuille or plane Couette flow
\citep{Tillmark,TrefethenSubcritical,Grossman}.
Later, sustained turbulence was seen
in fluid simulations of several plasma systems with no absolute linear instabilities
\citep{Waltz, Scott1, Scott2, Drake}.
More recently,
turbulence in the absence of linear instability has been seen in kinetic plasma simulations involving
sheared equilibrium flow \citep{EdmundPRL, BarnesRotation}.
These examples demonstrate that linear eigenmode analysis can give
deceiving predictions for nonlinear behaviour,
a conclusion supported by other studies of plasma turbulence in which
linear instability exists, but
the fluctuations in saturated nonlinear turbulence are different in character from the linear
eigenmodes \citep{Biskamp,Friedman1,Friedman2}.

For the cases of turbulence without absolute linear instability in the presence of sheared flows,
for both neutral fluids and plasmas, the phenomenon of transient linear amplification
appears to be important
\citep{TrefethenSubcritical, FarrellIoannou, Grossman, DelSoleSurvey, TrefethenEmbree, Schmid, Newton, Alex}.
Also known as non-modal amplification, transient amplification is a period of growth before
the exponential decay sets in.
A necessary (but not sufficient) condition for transient amplification
is that the linear operator must not have a complete orthogonal set of eigenvectors.
Analysis of transient amplification has also been applied to plasma systems in which
linear instability exists but transient linear behaviour may be
more important \citep{Camargo, Camporeale2010, Friedman2, SquirePRL, SquireApJ}.

In the present work, we consider the phenomena of transient amplification and turbulence
without instability for the case of the weakly-collisional
plasma drift wave driven by a density gradient \citep{Galeev,Krall3},
neglecting toroidal effects.
This oscillation has been called the ``universal instability'' as it was at first
thought to be linearly unstable in any plasma with a density gradient.
The oscillation is electrostatic in nature, and linear instability can indeed be found
in slab geometry without magnetic shear.
It was later established that when any amount of magnetic shear is included in the model,
no matter how small,
the eigenmodes with $k_y \rho_i \ll 1$ become linearly damped,
where $k_y$ is the perpendicular wavenumber and $\rho_i$ is the ion gyroradius
\citep{Ross, Tsang, Antonsen}.
However, instability can persist at $k_y \rho_i \ge 1$ when the collisionality is
very low \citep{usUniversalInstability}.
When both magnetic shear and modest collisions are included,
the mode becomes linearly stable at all $k_y \rho_i$.
It is this latter situation that we will consider here.
(As an aside, the small-but-finite-shear eigenmodes
are different from zero-shear modes not only for the
universal instability, but also for the slab ion-temperature-gradient (ITG) instability.
Specifically, in \citet{Plunk2014}, considering the slab mode by setting
curvature drift effects $\omega_d$ to 0, the small-shear limit of (19) is different from
the zero shear result (12).)

Fluid versions of this drift wave system were investigated by \citet{Scott1,Scott2}
and \citet{Drake}.
In these models, all linear modes were damped, yet nonlinearly, turbulence could be sustained.
In \citet{Drake} it was pointed out that an important secondary
instability in this system is a drift wave rotated by $\pi/2$ in the plane
perpendicular to the magnetic field, since the stabilizing effects of magnetic
shear are reduced in this orientation.
In this rotated drift instability, the role of the equilibrium density gradient is replaced in a sense by the
perturbed density gradient, the latter represented in the nonlinear term.
However, since the nonlinearity is conservative, the rotated drift wave does not inject energy
into the system in the way that can be done by a perturbation feeding off the density gradient of the equilibrium.
Several authors have also investigated transient linear effects in fluid drift wave models
\citep{Camargo, Friedman1, Friedman2}. However, these studies did not
include magnetic shear, which we will find to have a very strong effect on
some aspects of the transient behaviour.

In the following sections, we establish three results concerning the drift wave system in slab geometry
with magnetic shear,
extending previous work by considering a kinetic description and exploring the role of transient linear amplification.
For these studies, the initial-value gyrokinetic code gs2 \citep{gs2} will be used
for numerical computations.
First, in section \ref{sec:transient} we demonstrate that
significant transient linear amplification is possible in this system.
The transiently growing fluctuations
resemble the zero-shear eigenmodes
in the dependence of their instantaneous growth rate on wavenumber,
though the transients are sensitive to magnetic shear
in their global time-dependence.
Second, in section \ref{sec:turbulence} we demonstrate that universal-mode turbulence without linear
instability, seen previously in fluid models  \citep{Scott1,Scott2,Drake},
is also possible in the gyrokinetic model.
Finally, we prove there is a relationship between
transient linear amplification and nonlinear turbulence in this system,
using similar arguments to the proof for hydrodynamics in \citet{DelSoleNecessity}
and arriving at similar conclusions.
Specifically, if the physical parameters are such that turbulence can be sustained,
typical states of the turbulence must be amplified linearly for some amount of time, even if these states
eventually decay exponentially.
Two more precise statements of this result and their proof will be given in section \ref{sec:proof}.

@@

We consider slab geometry with magnetic shear, so the magnetic field in Cartesian
coordinates $(x',y',z')$ is $\vect{B} = [\vect{e}_{z'} + (x' / L_s) \vect{e}_{y'} ]B$
for some constant $B$ and shear length $L_s$.  A field-aligned coordinate system
is introduced: $x = x', \;y = y' - x' z' / L_s, \; z = z'$,
so $\vect{B}\cdot\nabla x = \vect{B}\cdot\nabla y = 0$ and $\vect{B}\cdot\nabla z = B$.
In this geometry, we consider the nonlinear electrostatic
gyrokinetic equation \citep{FriemanChen},
dropping toroidal effects:
\begin{eqnarray}
\label{eq:gke}
\frac{\partial h_s}{\partial t}
+ v_{||} \frac{\partial h_s}{\partial z}
+ \frac{1}{B} \left[
\frac{\partial \left<\Phi\right>_{\vect{R}_s}}{\partial X_s} \frac{\partial h_s}{\partial Y_s}
-\frac{\partial \left<\Phi\right>_{\vect{R}_s}}{\partial Y_s} \frac{\partial h_s}{\partial X_s}
\right]
-\left< C_s\left\{h_s\right\} \right>_{\vect{R}_s} \\
=
\frac{q_s}{T_s} f_{Ms} \frac{\partial \left<\Phi\right>_{\vect{R}_s}}{\partial t}
-\frac{f_{Ms}}{B L_n} \frac{\partial \left< \Phi\right>_{\vect{R}_s}}{\partial Y_s}
\left[ 1 + \left( \energy_s-\frac{3}{2}\right) \eta_s\right],
\nonumber
\end{eqnarray}
together with the quasineutrality condition
\begin{equation}
-\frac{n e \Phi}{T_i} + \int d^3v \left<h_i\right>_{\vect{r}}
=
\frac{n e \Phi}{T_e} + \int d^3v \left<h_e\right>_{\vect{r}}.
\label{eq:qn}
\end{equation}
Here $s \in \left\{i,e\right\}$ subscripts denote species,
$\Phi(t,\vect{r})$ is the electrostatic potential, $h_s(t,\vect{R}_s,\energy_s,\mu)$ is the nonadiabatic distribution function,
$\energy_s = m_s v^2/(2 T_s)$, $\mu = v_{\bot}^2/(2B)$,
and $f_{Ms}$ is the leading-order Maxwellian.
The scale length of the equilibrium density $n$ is $L_n = -n/(dn/dx)$,
$q_i = e = -q_e$ is the proton charge,
$T_s$ is the species temperature,
$\eta_s=(d \ln T_s/dx)/(d \ln n/dx)$,
$C_s$ is the collision operator,
and
$\left<\right>_{\vect{r}}$
and
$\left<\right>_{\vect{R}_s}$
denote gyroaverages at fixed particle position $\vect{r}=(x,y,z)$ and guiding-center position $\vect{R}_s=(X_s,Y_s,Z_s)$.
The distinction between $z$ and $Z_s$ may be neglected.
We consider a flux tube domain, in which the fluctuating quantities can be taken
to be periodic in $x$ and $y$.
Expanding the fluctuating quantities in spatial Fourier modes as $\Phi = \sum_{k_x,k_y} \Phi_{k_x,k_y}(t,z) \exp(i k_x x + i k_y y)$
and $h_s = \sum_{k_x,k_y} h_{s,k_x,k_y}(t,z,\energy_s,\mu) \exp(i k_x X_s + i k_y Y_s)$,
we may replace the gyro-averages with Bessel functions:
$\left<\Phi_{k_x,k_y} \exp(i k_x x + i k_y y)\right>_{\vect{R}_s}
=J_{0s} \Phi_{k_x,k_y} \exp(i k_x X_s + i k_y Y_s)$
and
$\left<h_{s,k_x,k_y}\exp(i k_x X_s + i k_y Y_s)\right>_{\vect{r}} = J_{0s} h_{s,k_x,k_y}\exp(i k_x x + i k_y y)$
with $J_{0s} = J_0(k_{\perp} v_{\perp} / \Omega_s)$
and $\Omega_s = e B/m_s$.
The perpendicular wavenumber is given by
\begin{equation}
k_{\perp} = \sqrt{k_y^2 + (k_x - k_y z/L_s)^2}.
\label{eq:kperp}
\end{equation}
The system is uniquely specified by the following dimensionless parameters:
$L_s / L_n$, $\eta_i$, $\eta_e$, $T_e / T_i$, and collisionality $\nuee L_n / v_i$,
where $\nuee = \sqrt{2}\pi n e^4 \ln\Lambda/[(4\pi\epsilon_0)^2 m_e^{1/2} T_e^{3/2}]$ is the electron-electron collision frequency
and $v_i = \sqrt{2 T_i / m_i}$.

@@

In contrast to the agreement shown in figure
\ref{fig:comparingTransientTo0Shear} for the amplification phase
following the maximum growth rate, we do not find such close agreement
between the zero-shear dispersion relation and the finite-shear
dynamics for very early times, before the maximum growth rate occurs.
In this initial period, a single dominant wave packet has not yet
overtaken other fluctuations present in the initial random noise.  We
also do not see such close agreement between the zero-shear and
finite-shear behaviour during the decay phase of the latter, i.e. at
large times.  The finite-shear decay rate as $t \to \infty$
corresponds to an eigenvalue, and it was established in previous work
\citep{Ross,Tsang,Antonsen} that the finite-shear eigenvalues
are not continuously related to the zero-shear eigenvalues.  In an
eigenmode, points at all $z$ have a coherent time-dependence.  Thus,
this damping rate is determined by a coherent interaction between
points of different $k_\perp(z)=k_y\sqrt{1+(z/L_s)^2}$, and between
all $k_{||}$ Fourier components.  This damping rate is thus not well
approximated by the zero-shear dispersion relation which assumes a
single $k_\perp$ and single $k_{||}$.

@@

\bibliographystyle{jpp}
